Exploitation Techniques and Mitigations

### 2016-01-18 docs/paper.tex
Reorganize source code and sections

\usepackage{wrapfig}
\newminted{nasm}{
    fontsize=\scriptsize,
    frame=leftline,
    framesep=2mm,
    autogobble,
    linenos}

@@

Exploiting binaries was comparatively easy ten to fifteen years ago. There were
no special mitigation mechanisms in place denying even the most simplest
exploits. This is the point in time where we will start of. First we talk about
two very simple exploits, namely the Format String Exploit and the Buffer
Overflow in combination with Shell Code. Although there is a huge collection of
exploitation techniques known to the public, we will only look at a very small
fraction of them in this project.

@@

Note that these registers can be addressed partially allowing one to write only
to the lower \SI{16}{bit}, for example, as displayed in
\cref{fig:single_register} taken from \textit{null
programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
December of 2015}.

@@

The first exploitation technique we will discuss builds upon the interpretation
of format strings. \texttt{printf} is a C function of the standard library
which will interpret such strings and print them to \texttt{stdout}. As the
name already tells you, the supplied string contains \textit{formatter}
describing how to actually handle additional arguments. If you are unfamiliar
with \texttt{printf} please have a look at the man
page\footnote{\url{http://linux.die.net/man/3/printf}}.

Taking a closer look at \texttt{printf} we can see that its first argument is a
format string followed by a variable number of additional arguments. In C you
don't know how many arguments have been supplied when a function with a
variable number of arguments is called. Some instances work around this by
taking an argument count as their first argument, others expect you to
terminate with a special symbol (usually \texttt{NULL}). \texttt{printf} uses
the format string to derive how many arguments have been supplied. Calling
\texttt{printf}, for example, with the string \texttt{"\%d + \%d = \%d"}
assumes that (at least) three arguments have been provided.

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \cfile{../format_string/main.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.5\textwidth}
        \begin{code*}{linenos=false}
            > echo foobar | ./main
            You entered:
            foobar

@@

The exploit comes from the notion that a format string provided by an attacker
gets interpreted. The program shown in \cref{src:format_string_exploit} will
take an arbitrary string from \texttt{stdin} and pass it on to \texttt{printf}.
For simple inputs (not containing formatter) this works fine. But as soon as
formatter are provided, \texttt{printf} is going to access the locations where
the corresponding arguments \emph{would} be located. From the calling
convention described in \cref{sub:calling_convention} we know that these
arguments would be located on the stack, therefore \texttt{printf} will print
whatever lies on the stack.

To fully exploit the provided example, note that an attacker in this scenario
wants to get a hold of the hardcoded password stored in \texttt{passwd}. Since
local variables are placed on the stack \texttt{printf} will be able to read
the password if enough formatters are provided:

@@

Here we use Python to craft the format string for us. As we can see the
password is printed (ASCII encoded). Byte order is swapped because of
endianness (see \cref{sub:endianness}). Apart from the password we also gather
a bunch of pointers, these can be used later on to break ASLR (see
\cref{sub:info_leak}).

We would like to point the reader to the book \textit{Hacking: The Art of
Exploitation} \cite[pp.~167]{art_of_exploitation} for more details about this
technique. We will come back to this technique later on to show that
\texttt{printf} enables even more sophisticated attacks.

@@

The second type of exploits we'll look at is known as Buffer Overflows and as
on may already derive from the name, this is about submitting more data to a
buffer than it was originally designed for. This setup can be exploited when
bound checking is done wrong or not at all. An attacker is therefore able to
overwrite data (or instructions) next to the buffer's location.

@@

Lets start of by examining the stack holding a buffer as local variable, see
\cref{fig:stack_frame}. Right now the buffer contains the string
\texttt{"hello"} followed by a terminator. Since the buffer has been allocated
to hold a maximum of \SI{12}{\byte} this fits. If data is written to the buffer
larger than \SI{12}{\byte} the following variable (or parameter) \texttt{bar}
will be overwritten, followed by the saved frame pointer and the return
address. If even more data is supplied the following stack frame will be
overwritten in the same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he can therefore inject arbitrary (malicious) into the
stack frame. This could be, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward lets go beyond that and see what happens when
changing the return address.

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \cfile{../overflow/overflow.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.5\textwidth}
        \codefile{../overflow/objdump.txt}
    \end{minipage}
    \caption{Program vulnerable to buffer overflows}
    \label{src:buffer_overflow}
\end{listing}

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. Lets start this section by crafting some shell code.

\begin{nasmcode}
    xor     eax, eax    ;Clearing eax register
    push    eax         ;Pushing NULL bytes
    push    0x68732f2f  ;Pushing //sh
    push    0x6e69622f  ;Pushing /bin
    mov     ebx, esp    ;ebx now has address of /bin//sh
    push    eax         ;Pushing NULL byte
    mov     edx, esp    ;edx now has address of NULL byte
    push    ebx         ;Pushing address of /bin//sh
    mov     ecx, esp    ;ecx now has address of address
                        ;of /bin//sh byte
    mov     al, 11      ;syscall number of execve is 11
    int     0x80        ;Make the system call
\end{nasmcode}

This piece of assembly sets up the parameters for the \texttt{execve} system
call and than invokes to replace the currently running process with a shell.
\texttt{execve} takes three arugments, a string of the program to execute (here
\texttt{"/bin//sh"} + terminator), a list of arguments for that program and a
list of environment variables. Its system call number is 11 and it will accept
\texttt{NULL} for both lists. The double slash in the first argument is used to
prevent null bytes inside the shell code. The function which reads the shell
code may truncate it upon reading a null byte, therefore we have to work around
this without chaning the underlying semantics.

Running this code through an assembler yields binary code which can be placed
in the buffer. Finding the starting location of our buffer will be a little bit
more complicated, we cannot read it directly from the binary of the target
program so we'll examine it in a debugger.

Now we know that the buffer will be located at XXXXXXXX at runtime, but since
we got this address while running the program in a debugger it may be offset a
few bytes when run without debugger. This happens because environment variables
and meta information, like the program name, determin the stack starting
position (they are placed right after the stack). Hence we may not directly hit
the first instruction of our shell code right away, but since the buffer is
bigger than the actual payload we can improve our odds by prefixing the shell
code with \texttt{NOP} instructions. As long as the return address points
somewhere into this sequence of \texttt{NOP}s the CPU will \emph{slide} to the
next instruction. Therefore this is known as a \textit{\texttt{NOP} Sled}. We
append some arbitrary data to the shell code as offset to overwrite the return
address.

@@

\newpage

@@

>>>>

\usepackage{wrapfig}
\newminted{nasm}{
    fontsize=\scriptsize,
    frame=leftline,
    framesep=2mm,
    autogobble,
    linenos}
\newmintedfile{nasm}{
    fontsize=\scriptsize,
    frame=leftline,
    framesep=2mm,
    autogobble,
    linenos}

@@

Exploiting binaries was comparatively easy ten to fifteen years ago. There were
no special mitigation mechanisms in place denying even the most simplest
exploits. This is the point in time where we will start off. First we talk
about two very simple exploits, namely the Format String Exploit and the Buffer
Overflow in combination with Shell Code. Although there is a huge collection of
exploitation techniques known to the public, we will only look at a very small
fraction of them in this project.

@@

Note that these registers can be addressed partially allowing one to write only
to the lower \SI{16}{bit}, for example, as displayed in
\cref{fig:single_register} taken from \textit{null
programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
December 2015}.

@@

The first exploitation technique we will discuss builds upon the interpretation
of format strings. \texttt{printf} is a C function of the standard library
which will interpret such strings and print them to \texttt{stdout}. As the
name already tells you, the supplied string contains \textit{formatter}
describing how to handle additional arguments. If you are unfamiliar with
\texttt{printf} please have a look at the man
page\footnote{\url{http://linux.die.net/man/3/printf}} now.

Taking a closer look at \texttt{printf} we can see that its first argument is a
format string followed by a variable number of additional arguments.
\texttt{stdarg.h} describes a common implementation of this together with a
small example in their man
page\footnote{\url{http://linux.die.net/man/3/stdarg}}. As in that example,
\texttt{printf} trusts you that the number of arguments supplied is equal (or
greater) than the number of formatters. Calling \texttt{printf} with the format
string \texttt{"\%d + \%d = \%d"}, for instance, assumes that (at least) three
additional arguments are given.

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \cfile{../src/format_string/format.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.5\textwidth}
        \begin{code*}{linenos=false}
            > gcc -o format format.c

            > echo foobar | ./main
            You entered:
            foobar

@@

The exploit comes from the notion that a format string provided by an attacker
gets interpreted. The program shown in \cref{src:format_string_exploit} will
take an arbitrary string from \texttt{stdin} and pass it on to \texttt{printf}.
For simple inputs (not containing formatter) this works fine. But as soon as
formatter are provided, \texttt{printf} accesses the locations where the
corresponding arguments \emph{would} be located. From the calling convention
described in \cref{sub:calling_convention} we know that these arguments
\emph{would} be located on the stack, therefore \texttt{printf} will print
whatever lies on the stack at these positions instead.

An attacker in this scenario wants to get a hold of the hardcoded password
stored in \texttt{passwd}. Since local variables are placed on the stack
\texttt{printf} will be able to read the password if enough formatters are
provided:

@@

Here we use Python to craft the format string containing ten identifiers for
us. As we can see the password is printed (ASCII encoded). Byte order is
swapped because of endianness (see \cref{sub:endianness}). Apart from the
password we also gather a bunch of pointers, these can be used later on to
break ASLR (see \cref{sub:info_leak}).

We would like to point the reader to the book \textit{Hacking: The Art of
Exploitation}~\cite[pp.~167]{art_of_exploitation} for more details about this
and following techniques. We will come back to this technique later on to show
that \texttt{printf} enables even more sophisticated attacks.

@@

The second type of exploits we'll look at is known as Buffer Overflows and as
on may already derive from the name, this is about submitting more data to a
buffer than it was originally designed for. This setup can be exploited when
bound checking is done wrong or not at all. An attacker is therefore able to
overwrite memory behind the buffer's location.

\subsection{Disabling Mitigations}

The three mitigation mechanisms DEP, ASLR and Stack Cookies are enabled by
default nowadays and as mentioned in the introduction we start off at a pointer
where these mechanisms were not. So to run the provided examples we first have
to disable them. DEP and Stack Cookies can be disabled via compiler flags to
the extend necessary using \texttt{-z execstack} and
\texttt{-fno-stack-protector} respectively.

ASLR can be disabled globally so that \emph{new} process' has an unscrambled
memory layout:

\begin{code*}{linenos=false}
    > echo 0 > /proc/sys/kernel/randomize_va_space
\end{code*}

Writing \texttt{2} instead of \texttt{0} will switch ASLR back to its default
state. Root privileges are, of course, required for this. There is also another
way by invoking \texttt{setarch}:

\begin{code*}{linenos=false}
    > setarch `arch` -R ./binary
\end{code*}

\subsection{The Exploit}

@@

Lets start off by examining the stack containing a buffer \texttt{c} as local
variable, see \cref{fig:stack_frame}. Right now the buffer holds the string
\texttt{"hello"} followed by a terminator. Since it has been allocated
to hold a maximum of \SI{12}{\byte} this fits. If data is written to the buffer
larger than \SI{12}{\byte} the following variable (or parameter) \texttt{bar}
will be overwritten, followed by the saved frame pointer and the return
address. If even more data is supplied the following stack frame will be
overwritten in the same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he is able to inject arbitrary (malicious) code into
the stack frame. This could be, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward lets go beyond that and see what happens when
changing the return address.

\begin{listing}[h!]
    \begin{minipage}[t]{0.37\textwidth}
        \cfile{../src/buffer_overflow/overflow.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.63\textwidth}
        \codefile{../src/buffer_overflow/overflow_objdump.txt}
    \end{minipage}
    \caption{Program vulnerable to buffer overflows}
    \label{src:buffer_overflow}
\end{listing}

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. This example is taken from Dhaval Kapil's
blog\footnote{\url{https://dhavalkapil.com/blogs/Shellcode-Injection/} on
December 2015} there is also a section about this in \textit{Hacking: The Art
of Exploitation}~\cite[pp.~281]{art_of_exploitation}.

\subsection{Crafting Shell Code}

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \nasmfile{../src/shell_code/shellcode.asm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.5\textwidth}
        \codefile{../src/shell_code/shellcode_objdump.txt}
    \end{minipage}
    \caption{Assembly code opening up a shell upon execution}
    \label{src:shell_code}
\end{listing}

The piece of assembly shown in \cref{src:shell_code} sets up the parameters for
the \texttt{execve} system call and than invokes it to replace the currently
running process with a shell. \texttt{execve} takes three arguments, a string
of the program to execute (here \texttt{"/bin//sh"} + terminator), a list of
arguments for that program and a list of environment variables. Its system call
number is 11 and it will accept \texttt{NULL} for both lists. The double slash
in the first argument is used to prevent null bytes inside the shell code. The
function which reads the shell code may truncate it upon reading a null byte,
therefore we have to work around this without changing the underlying
semantics.

Running this code through an assembler yields binary code, shown in
\cref{src:shell_code}, which will be part of the payload.

\begin{code*}{linenos=false}
    \x31\xC0\x50\x68\x2F\x2F\x73\x68\x68\x2F\x62\x69\x6E\x89\xE3\x50\x89\xE2\x53\x89\xE1\xB0\x0B\xCD\x80
\end{code*}

\subsection{Examining the Target Binary}

Finding the starting location of our buffer will be a little bit more
complicated, we cannot read it directly from the binary of the target program
so, in \cref{src:shell_code2}, we'll examine it in a debugger .

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \cfile{../src/shell_code/vuln.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.5\textwidth}
        \codefile{../src/shell_code/vuln_gdb.txt}
    \end{minipage}
    \caption{Examining the target binary in \texttt{gdb}}
    \label{src:shell_code2}
\end{listing}

Now we know that the buffer will be located at \texttt{0xbffff4bc} (saved base
pointer will be at \texttt{0xbffff528}) at runtime, but it may be offset a few
bytes when run without debugger. This happens because environment variables and
meta information, like the program name, determine the stack starting position
(stack is placed right before environment variables). Hence we may not hit the
first instruction of our shell code right away, but since the buffer is bigger
than the actual payload we can improve our odds by prefixing the shell code
with \texttt{NOP} instructions. As long as the return address points somewhere
into this sequence of \texttt{NOP}s the CPU will \emph{slide} to the first
instruction of the shell code. Therefore this is known as a
\textit{\texttt{NOP} Sled}. We append some arbitrary data to the shell code as
offset to overwrite the return address. This is also illustrated in
\cref{fig:shell_code} where \textit{target} is new return address supplied by
the attacker. Using the maximum amount of \texttt{NOP}s possible would also be
a viable option, here we just went with the original example.

@@

Lets first calculate the distance between the start of the buffer and the
return address. The return address will be located \SI{4}{\byte} after the
saved base pointer location.

\[(\texttt{0xbffff528} + 4) - \texttt{0xbffff4bc} = 112\]

We prefix our shell code with a \texttt{NOP} Sled consisting of \SI{40}{\byte}
(opcode for \texttt{NOP} is \texttt{0x90}). Since our shell code is \SI{25}{\byte}
long we add 47 \texttt{'A'}s to gap the remaining distance. Lastly we have to add
the new return address which should point to the \texttt{NOP} Sled's center.

\[\texttt{0xbffff4bc} + 20 = \texttt{0xbffff4d0}\]

\subsection{Gimme that Shell already}

\codefile[linenos=false]{../src/shell_code/run.txt}

After injecting the payload we get a few lines of garbage and receive a prompt
by hitting return a few times. You can enter commands and receive output like
usual.

@@

\subsection{Info Leak}
\label{sub:info_leak}
